\documentclass[10pt,a4paper,oneside]{book}

\title{Lesson 2}
\author{}
\date{}

\input{../../../Templates/jointPreamble.tex}
\graphicspath{{figures/}}
\usepackage{subfiles}

\begin{document}
\maketitle

There are some special sets of numbers.

There is the set of positive integers
\begin{center}
    $\{1,2,3,4,\dots\}$
\end{center}

However, we cannot ignore $0$ or the negative numbers, so we have $\mathbb{Z}$, the entire set of integers.
\begin{center}
    $\mathbb{Z}=\{\dots,-4,-3,-2,-1,0,1,2,3,4,\dots\}$
\end{center}

Addition in $\mathbb{Z}$ gives some nice axioms
\begin{itemize}
    \item \textbf{commutativity}: for any $a,b\in \mathbb{Z}$, $a+b=b+a$
    \item \textbf{associativity}: for any $a,b,c \in \mathbb{Z}$, $(a+b)+c=a+(b+c)$
    \item $\mathbb{Z}$ has an additive identity element, namely $0$, for any $a\in \mathbb{Z}, a+0=0+a=a$
    \item $\mathbb{Z}$ has additive inverses: for any $a\in \mathbb{Z}$, there exists $b\in \mathbb{Z}$ such that $a+b=b+a=0$. (Of course, $b=-a$)
\end{itemize}

We know that $\mathbb{Z}$ has multiplication that is also commutative, associative, and has an identity element, $1$.

We also have that 
\begin{center}
    for any $a,b,c\in \mathbb{Z}$, $a(b+c)=(ab)+(ac)$
\end{center}

We move to the rational numbers due to not having multiplicative inverses.
\begin{center}
    $\mathbb{Q}=\{\frac{m}{n}|m,b\in\mathbb{Z} \text{ and } n\neq 0\}$
\end{center}

The set $\mathbb{Q}$ together with the operations of addition and multiplication is what is called a field, and these operation satisfy all the nice properties we want them to have. However there is something missing from $\mathbb{Q}$.

For example there is no number $x\in \mathbb{Q}$ such that $x^2=2$, since, yes in $\mathbb{R}$ we have a solution of $\sqrt{2}$, but in $\mathbb{Q}$ this does not exist.

\begin{definition}
    Let $S$ be a subset of $\mathbb{R}$.

    \begin{itemize}
        \item A number $x\in \mathbb{R}$ is called an upper bound for $S$ if for all $y\in S$, we have $y\leq x$. If an upper bound for $S$ exists, we say that $S$ is bounded above.
        \item A number $x\in \mathbb{R}$ is called a lower bound for $S$ if for all $y\in S$, we have $y\geq x$. If a lower bound for $S$ exists, we say that $S$ is bounded below.
        \item If $S$ has an upper bound and a lower bound, we say that $S$ is bounded.
    \end{itemize}
\end{definition}

There are some special bounds.
\begin{definition}
    Let $S$ be a subset of $\mathbb{R}$.

    \begin{itemize}
        \item A number $x\in \mathbb{R}$ is called a least upper bound of $S$ if $x$ is an upper bound for $S$ and if for every $z\in \mathbb{R}$ such that $z$ is an upper bound of $S$, we have $x\leq z$.
        \item A number $x\in \mathbb{R}$ is called a greatest lower bound of $S$ if $x$ is a lower bound for $S$ and if for every $z\in \mathbb{R}$ such that $z$ is a lower bound of $S$, we have $x\geq z$.
    \end{itemize}
\end{definition}

In other words, an upper bound of a subset $S\subseteq \mathbb{R}$ is any number greater than or equal to all of the numbers in $S$.
If we can find a smallest possible upper bound for $S$, then this number is the least upper bound of $S$. Similarly, a lower bound of $S$ is a number that is less than or equal to all of the numbers in $S$.
If we can find a greatest possible lower bound for $S$, then this number is the greatest lower bound of $S$.

The most important subsets of $\mathbb{R}$ that we will be regularly using are intervals:
\begin{definition}
    A subset $I$ of $\mathbb{R}$ is called an interval if, for any $a,b \in I$ and $x\in \mathbb{R}$ such that $a\leq x\leq b$, we have $x\in I$.
\end{definition}

Informally, an interval $I$ is a subset of $\mathbb{R}$ without any ``holes''. If $a$ and $b$ are in $I$, then all the numbers between $a$ and $b$ are also in $I$. There are two types of intervals commonly used.
\begin{definition}
    Let $a\leq b$ be any two real numbers. The open interval $(a,b)$ is defined as the set 
    \[ (a,b)=\{x\in\mathbb{R}|a<x<b\} \]
    The closed interval $[a,b]$ is defined as the set
    \[ [a,b] = \{x\in \mathbb{R}|a\leq x\leq b\}\] 
\end{definition}

There are also half-open intervals.

We have one more definition
\begin{definition}
    Let $S$ be a subset of $\mathbb{R}$. If $\text{sup} S\in S$, we say that $\text{sup} S$ is the maximal element of $S$ (or simply the maximum of $S$). 
    If $\text{inf} \in S$, we say that $\text{inf} S$ is the minimal element of $S$ (or simply the minimum of $S$).
\end{definition}


\subsubsection*{Homework}

\begin{enumerate}
    \item Show that there is no number $x\in \mathbb{Q}$ such that $x^2=2$.
    \item 
    \begin{itemize}
        \item Show that the intersection of any two intervals is an interval.
        \item Show that the union of any two intervals need not be an interval.
    \end{itemize}
    \item If $A,B$ are bounded intervals with $A\cap B\neq \emptyset$, then show that $\text{sup}(A\cap B)=\min\{\text{sup}A,\text{sup}B\}$.
\end{enumerate}


\end{document}