\documentclass[../maths.tex]{subfiles}
\graphicspath{{\subfix{../figures/}}}
\begin{document}
\chapter{Algebra \& Functions}
\section{Indices}
\subsection*{Subsets of Real Numbers}
\subsubsection*{Introducing Subsets of Real Numbers}
Natural numbers are represented by $\mathbb{N}$. They are just the counting numbers - like $1,2,3,4,5,6,\dots$. This does not include 0 or negative numbers.

Integers are represented by $\mathbb{Z}$. This includes all the natural numbers and also includes $0,-1,-2,-3,\dots$. It is twice the size of natural numbers plus a zero.

Rational numbers are represented by $\mathbb{Q}$. This would include $\frac{1}{2}, \frac{2}{3}, -\frac{3}{4}, -\frac{5}{7}, -\frac{7}{2}$ along with the natural numbers and integers.

The real numbers are represented by $\mathbb{R}$. This includes everything above, along with things such as $\sqrt{2}, \sqrt{3}, \pi, e$.

The complex numbers are based on if we allowed to square root $-1$. We define this as $i$. The complex numbers will include things such as $2i, 3+i$.
\subsection*{The Laws of Indices}
\subsubsection*{The Laws of Indices}

We should know that $x^2 = x \times x$, and $x^3 = x \times x \times x$. The index tells us how many times we are multiplying $x$ by itself.

When we put the $x$ as $x^2$, we can see that $x^2\times x^2 = x\times x\times x\times x = x^4$ or $(x^2)^2$.

As we can see, when multiplying $x^p \times x^q = x^{p+q}$. 

Also when we have $(x^p)^q = x^{pq}$. Of course we know that $pq = qp$, and ad can also see that $(x^q)^p = (x^p)^q$.

Now let's imagine what we have $x^5 \div x^3 = \frac{x \times x\times x\times x\times x}{x\times x\times x} = x\times x = x^2$.

When we are dividing, then $x^p \div x^q = x^{p-q}$.

Let's say we have $x^{3.5}$. As long as the power is a rational number (in this case $3.5=\frac{7}{2}$), then we can have an idea on what it is. We can write $x^{\frac{7}{2}}$ as $x^{\frac{1}{2}\times 7}$.
This is the same now as $(x^{\frac{1}{2}})^7$.

This shows us our next rule - $x^{\frac{1}{p}} = \sqrt[p]{x}$.

So the above equation can be written as $(\sqrt{x})^7$.

Now let's also consider $x^0$. If you think about writing this as $x^{2-2}$, this equals $\frac{x^2}{x^2} = 1$.

Therefore, $x^0=1$.

Now we can look at $x^{-1} = x^{4-5} = \frac{x^4}{x^5}$. So from this we get $\frac{x\times x\times x\times x}{x\times x\times x\times x\times x} = \frac{1}{x}$.

This means that $x^{-1} = \frac{1}{x}$.

We have the rule then that $x^{-p} = \frac{1}{x^p}$.

\subsubsection*{Examples of Negative Indices}
\ex $2^{-3} = $

\ex $3^{-4} = $

\ex $5^{-2} = $

\ex $\left(\frac{1}{4}\right)^{-2} = $

\ex $\left(\frac{2}{3}\right)^{-3} = $

\subsubsection*{Examples of Positive Rational Indices}

\ex $36{\frac{1}{2}} = $

\ex $81^{\frac{1}{4}} = $

\ex $\left(\frac{1}{8}\right)^{\frac{1}{3}} = $

\ex $25^{\frac{3}{2}} = $

\ex $\left(\frac{8}{27}\right)^{\frac{2}{3}} = $

\subsubsection*{Examples of Negative Rational Indices}

\ex $8^{-\frac{1}{3}} = $

\ex $16^{-\frac{3}{4}} = $

\ex $4^{-\frac{5}{2}} = $

\ex $\left(\frac{36}{49}\right)^{-\frac{1}{2}} = $

\ex $\left(\frac{10000}{16}\right)^{-\frac{5}{4}} = $

\subsubsection*{More Complicated Examples}

\begin{example}
    $2^3\times 8^{-\frac{5}{3}} \times \frac{1}{\sqrt{2}} = 2^k$. Find $k$.

    For this problem, you want to write everything in terms of $2$ to the power of something. We can rewrite this equation as 
    \[ 2^3 \times (2^3)^{-\frac{5}{3}} \times 2^{-\frac{1}{2}} \]

    So this can be rewritten as $2^3\times 2^{-5} \times 2^{-\frac{1}{2}}$, and using laws of indices, we can see that this is equivalent to 
    $2^{-\frac{5}{2}}$. So $k = -\frac{5}{2}$.
\end{example}
\ex Write $\frac{x^2y^5}{\sqrt{x}}\div \frac{x^{\frac{3}{2}}}{y^7}$ as a product of powers of $x$ and $y$.

\subsubsection*{Examples of Simplifying Expressions}

\ex $5a^3b^2c \times 6a^8bc^{-3} = $

\ex $(60a^4b^2c) \div (12a^8b^5c^{-4}) = $

\ex $\frac{(3x)^3\times (2x^3)^4}{(6x^8)^2} = $

\subsubsection*{Write in the form of $2^k$}
\smallbreak
\begin{example}
    Write $\frac{\sqrt{2}}{4^3}$ in the form $2^k$.

    We can rewrite $\sqrt{2} = 2^{\frac{1}{2}}$ and $4^3 = (2^2)^3 = 2^6$. So now we have $\frac{2^{\frac{1}{2}}}{2^6} = 2^{-\frac{11}{2}}$.
\end{example}
\ex Write $8^4 \times \frac{2}{\sqrt[3]{16}}$ in terms of $2^k$.

\subsubsection*{Write in the form of $3^k$}
\smallbreak
\begin{example}
    Write $\sqrt[3]{3} \times \sqrt[3]{9}$ in terms of $3^k$.

    We can rewrite this as $3^{\frac{1}{3}} \times (3^2)^{\frac{1}{3}} = 3^{\frac{1}{3}}\times 3^{\frac{2}{3}} = 3^1$.
\end{example}

\ex Write $\frac{\sqrt[5]{27}}{\sqrt{3}}\times 81$ in terms of $3^k$.

\subsubsection*{Write in the form of $4^k$}
\smallbreak
\begin{example}
    Write $\frac{16}{\sqrt[4]{5}}$ in terms of $4^k$.

    This can be rewritten as $\frac{4^2}{4^{\frac{1}{5}}}$, so this is equivalent to $4^{\frac{9}{5}}$.
\end{example}
\ex Rewrite $2\times \sqrt[3]{16} \times \sqrt[5]{64}$ in terms of $4^k$.

\subsubsection*{Write in the form $5^k$}
\smallbreak
\begin{example}
    Rewrite $\frac{125}{\sqrt[3]{25}}\times \sqrt{5}$ in terms of $5^k$.

    We first start off with $\frac{5^3}{(5^2)^{\frac{1}{3}}}\times 5^{\frac{1}{2}}$. 

    This is equal to $5^{\frac{7}{3}}\times 5^{\frac{1}{2}} = 5^{\frac{17}{6}}$.
\end{example}
\ex Rewrite $\frac{\sqrt[3]{50}}{\sqrt{625}}\times \sqrt[3]{12.5}$ in terms of $5^k$.

\section{Surds}
\subsection*{Simplifying Surds}
\subsubsection*{Introducing Surds and Simplifying Surds}
You can quite easily build up a list of what you believe are surds - $\sqrt{1}=1$, $\sqrt{2}$ and $\sqrt{3}$ are surds, $\sqrt{4}=2$.
If the number under the square root if a square number, then obviously it will not be a surd.

Let's get an example with $\sqrt{8}$. Using our indices knowledge, we can write this as $8^{\frac{1}{2}}=(4\times 2)^{\frac{1}{2}}=4^{\frac{1}{2}}\times 2^{\frac{1}{2}}$.

This can be written as $\sqrt{4}\times \sqrt{2}=2\sqrt{2}$.

Now let's look at $\sqrt{12}$. Now we can write this as $\sqrt{6}\times \sqrt{2}$, but there is no real point in doing this, since none of these can be simplified.
We are looking for any square numbers that can go in 12. So we can write this as $\sqrt{4}\times \sqrt{3}$ which is equivalent to $2\sqrt{3}$.

This leads us to the rule $\sqrt{ab}=\sqrt{a}\times \sqrt{b}$.
\subsubsection*{Simplifying Surds Examples}
\ex $\sqrt{18}$

\ex $\sqrt{200}$

\ex $\sqrt{48}$

\ex $\frac{\sqrt{12}}{\sqrt{300}}$

\ex $\sqrt{24}\times \sqrt{150}$

\subsubsection*{Adding/Subtracting Surds}
Let's start with an example.

If we are given $\sqrt{20}+\sqrt{180}$, we cannot add the two together. In most cases $\sqrt{a}+\sqrt{b}\neq \sqrt{a+b}$.

We have $\sqrt{4}\times \sqrt{5} + \sqrt{36} \times \sqrt{5}$ when we simplify this. Now we can simplify this as $2\sqrt{5}+6\sqrt{5}$. 
Surds can be combined like `like' terms in algebra. So the answer so this expression is $8\sqrt{5}$.

Here's two more examples to try with the answer given.

\textbf{Example 1: } $\sqrt{63}-\sqrt{28}=\sqrt{7}$

\textbf{Example 2: } $\sqrt{108}+\sqrt{72} = \sqrt{2}+\sqrt{3}+\sqrt{5}$

\begin{example}
    $\sqrt{3}(\sqrt{2}+5)$.

    Expanding this, we get that $\sqrt{6}+5\sqrt{3}$. Note that if you can split up a surd, you can also do the reverse and multiply them back together.
\end{example}

\ex $6(\sqrt{3}+\sqrt{6})$

\ex $\sqrt{5}(8-\sqrt{7})$

\ex $\sqrt{6}(\sqrt{15}-2\sqrt{2})$

\ex $\sqrt{12}(\sqrt{50}+3\sqrt{10})$

\begin{example}
    $(2+\sqrt{2})(3-\sqrt{5})$. This example is simliar to above, but with double brackets instead.

    So using a technique of your choice, you should end up with $6-2\sqrt{5}+3\sqrt{2}-\sqrt{10}$.
\end{example}

\ex $(2-\sqrt{5})(2+\sqrt{5})$

\ex $(3+\sqrt{2})(2+\sqrt{3})$

\ex $(\sqrt{2}+1)(\sqrt{3}-\sqrt{5})$

\ex $(2\sqrt{3}+3\sqrt{5})(2\sqrt{2}-5\sqrt{3})$
\subsection*{Rationalising the Denominator}
In general, we do not want a surd in the denominator of a fraction such as $\frac{1}{\sqrt{2}}$. We can use a technique known as rationalising the denominator. In order to do this, we want to multiply by $1 = \frac{\sqrt{2}}{\sqrt{2}}$. So when we multiply $\frac{1}{\sqrt{2}}$ by this, we get that the answer is $\sqrt{2}{2}$.

\ex $\frac{2}{\sqrt{3}}$

\ex $\frac{10}{\sqrt{5}}$

\ex $\frac{9}{2\sqrt{3}}$

\begin{example}
    Let's try rationalising $\frac{1}{1+\sqrt{2}}$

    For this, we must use the identity $x^2-y^2=(x-y)(x+y)$. If we multiply $(1+\sqrt{2})$ by $(1-\sqrt{2})$, we are able to rationalise the denominator in this case.

    So multiplying by $\frac{1-\sqrt{2}}{1-\sqrt{2}}$ all the way simplifies to $-1+\sqrt{2}$, which is the rationalised form of what was given.
\end{example}

\ex $\frac{2}{\sqrt{2}+2}$

\ex $\frac{3}{4-\sqrt{5}}$

\ex $\frac{1+\sqrt{2}}{3-\sqrt{2}}$

\ex $\frac{4+2\sqrt{3}}{3+3\sqrt{2}}$

Challenge Problems 
\begin{enumerate}
    \item \begin{tikzpicture}
        % Define points
        \coordinate (B) at (0,0);
        \coordinate (C) at (5,0);
        \coordinate (A) at (2.5,3);
        
        % Draw triangle
        \draw (B) -- (A) -- (C) -- cycle;
        
        % Label points
        \node[above] at (A) {A};
        \node[left] at (B) {B};
        \node[right] at (C) {C};
        
        % Label side
        \node[below] at ($(B)!0.5!(C)$) {$6\sqrt{2} - 2\sqrt{5}$};
    \end{tikzpicture}
    
    $\triangle ABC$ has area $5$. Find the exact perpendicular height of the triangle.

    \item Rationalise the denominator of $\frac{1}{\sqrt{2}+\sqrt{3}-\sqrt{5}}$
\end{enumerate}

\subsection*{Problem Solving}
\section{Quadratics}
\subsection*{The Difference of Two Squares}
\subsection*{Factorising Quadratics}
\subsection*{Sketching Quadratics from Factorised Form}
\subsection*{Completing the Square}
\subsection*{Sketching Quadratics from Completed Square Form}
\subsection*{Solving Quadratics}
\subsection*{Using the Discriminant}
\subsection*{Using the Quadratic Formula}
\subsection*{Sketching Quadratics Using the Quadratic Formula}
\subsection*{Sketching Quadratic Using a Calculator}
\subsection*{Using Quadratic Methods for Solving}
\section{Simultaneous Equations}
\subsection*{The Elimination Method}
\subsection*{The Substitution Method}
\subsection*{Further Simultaneous Equations}
\section{Inequalities}
\subsection*{Introducing Inequalities, Set Notation and Interval Notation}
\subsection*{Linear Inequalities}
\subsection*{Quadratic Inequalities}
\subsection*{Discriminant Inequalities}
\subsection*{More Inequalities}
\subsection*{Double and Triple Inequalities}
\subsection*{Representing Inequalities Graphically}
\section{Polynomials \& Rational Expressions}
\subsection*{Introducing Polynomials}
\subsection*{Polynomial Division}
\subsection*{The Factor Theorem}
\subsection*{Simplifying Algebraic Fractions}
\subsection*{Adding and Subtracting Algebraic Fractions}
\subsection*{Simplifying using Polynomial Division}
\section{Graphs \& Proportion}
\subsection*{Sketching Polynomials}
\subsection*{Sketching Cubics from Factorised Form}
\subsection*{Sketching Quartics from Factorised Form}
\subsection*{The Modulus Function}
\subsection*{Reciprocal Graphs}
\subsection*{Finding Points of Intersection}
\subsection*{Direct and Inverse Proportion}
\section{Functions}
\subsection*{Introducing Functions}
\subsection*{The Domain and Range of a Function}
\subsection*{One-to-One, Many-to-One, One-to-Many, Many-to-Many}
\subsection*{Restricting the Domain}
\subsection*{Even \& Odd Functions}
\subsection*{Composite Functions}
\subsection*{Inverse Functions}
\subsection*{Set Notation}
\subsection*{Interval Notation}
\section{Graph Transformations}
\subsection*{Introducing Transformations}
\subsection*{Translations}
\subsection*{Stretches}
\subsection*{Reflections}
\subsection*{Examples of Transformations}
\subsection*{Combining Transformations}
\section{Algebraic Fractions}
\subsection*{Algebraic Fractions}
\subsection*{Partial Fractions}
\subsection*{Partial Fractions: Repeated Factors}
\subsection*{Partial Fractions: Extensions}
\subsection*{Partial Fractions: with Binomial Expansion \& Integration}
\section{Modelling}
\end{document}