\documentclass[../satmath.tex]{subfiles}
\graphicspath{{\subfix{../figures/}}}
\begin{document}
\chapter{Non-Linear Equations and Functions}
\begin{enumerate}[label=\bfseries\arabic*.]

\item If $x^2-y^2=322$ and $x+y=23$, what is the value of $x-y$?

\item  
\[2x^2+5x+1=0\]
In the equation above, what is the largest possible value of $x$?

$\textbf{(A) } \frac{-5-\sqrt{17}}{4} \qquad \textbf{(B) } \frac{-2+\sqrt{17}}{5} \qquad \textbf{(C) } \frac{-2-\sqrt{17}}{5} \qquad \textbf{(D) } \frac{-5+\sqrt{17}}{4}$
 
\item If the equation $y=(x-3)(x-7)$ is graphed on the $xy$-plane, which of the following is an equivalent form that identifies the $x$-coordinate of the parabola's vertex as a constant?

$\textbf{(A) } y=(x-5)^2-4 \qquad \textbf{(B) } y=(x+5)^2-4 \qquad \textbf{(C) } y=(x-5)^2+21 \qquad \textbf{(D) } y=(x+5)^2+21$

\item How many real solutions exist for the equation $4x^2+16x+16=0$?
 
\item The function $f$ is defined as $f(x)=x^3-2x^2+4x+d$. If $(3,0)$ is a solution, what is the value of $d$?
 
\item If $m+n = 12$ and $m^2-n^2=48$, what does $m-n$ equal?
 
\item What is the sum of the solutions to the equation $8x^2-24x+100=0$?
 
\item In the $xy$-plane, the graph of $y=x^3-x^2-kx-24$ has the solutions $(e,0)$, $(f,0)$, and $(6,0)$. What is the value of $k$?
 
\item If $x^2-6x-27=0$ and $x>0$, what is the value of $x$?
 
\item If $f(x)=3x^2-5x+2$, which of the following is equal to $f(-2)$?

$\textbf{(A) } 0 \qquad \textbf{(B) } 4 \qquad \textbf{(C) } 12 \qquad \textbf{(D) } 24$
 
\item The function $f$ is defined as $f(x)=10\sqrt{x}$. For what value of $x$ does $f(x)=60$?
 
\item When the function $f$ is graphed in the $xy$-plane, the point $(3,8)$ lies on the graph. The function $g$ is defined as $g(x)=-f(x+2)-3$.
Which of the following points must lie on the graph of $g$ in the $xy$-plane?

$\textbf{(A) } (-3,9) \qquad \textbf{(B) } (-1,-5) \qquad \textbf{(C) } (1,-11) \qquad \textbf{(D) } (1,5)$

\item 
\begin{align*}
-6x+y=14\\
2x^2=y+6
\end{align*}

The graphs of the equations in the given system of equations intersect at the point $(x,y)$ in the $xy$-plane. What is one possible value of $x$?

$\textbf{(A) } -6 \qquad \textbf{(B) } -2 \qquad \textbf{(C) } 4 \qquad \textbf{(D) } 8$
 
\item The function $f$ is defined as $f(x)=6x^2$. Which of the following is a possible value of $a$ for which $f(a)=54$?

$\textbf{(A) } -5 \qquad \textbf{(B) } -3 \qquad \textbf{(C) } 2 \qquad \textbf{(D)} 9$
 
\item The function $f$ is defined as $f(x)=2x-5$ and the function $g$ is defined as $g(x)=-f(x)+3$. What is the value of $g(-4)$?

\item
\begin{align*}
x^2+y^2=333\\
-6x=y
\end{align*}
What is a possible value of $y$ in the system of equations above?

$\textbf{(A) } -9 \qquad \textbf{(B) } -3 \qquad \textbf{(C) } 3 \qquad \textbf{(D) } 18$
 
\item The function $f$ is defined as $f(x)=10x-15$. What is the value of $f(4)$?

\item If the quadratic equation $x^2-kx = 24$ has only integer solutions, which of the following cannot be a possible value of $k$?\

$\textbf{(A) } 11 \qquad \textbf{(B) } 10 \qquad \textbf{(C) } 5 \qquad \textbf{(D) } 2$
 
\item How many real solutions exist for the equation $2x^2-10x+5=0$?

\item If $a+b=12$ and $a-b=8$, what is the value of $a^2-b^2$?
 
\item The equation of a parabola graphed in the $xy$-plane is given as $y=x^2-8x+12$. Which of the following is an equivalent form that displays the 
solutions as constants or coefficients?

$\textbf{(A) } y=(x-2)(x-6) \qquad \textbf{(B) } y=(x+2)(x+6) \qquad \textbf{(C) } y=(x-4)^2-4 \qquad \textbf{(D) } y=(x-4)^2+12$
 
\item In the $xy$-plane, the point $(1,5)$ lies on the graph of $f(x)=x^3+3x^2+cx-12$. What is the value of $c$?

\item If the solutions to the equation $2x^2-13x+15=0$ are represented as $c$ and $d$, what is the value of $c+d$?

\item The equation of a parabola in the $xy$-plane is given as $y=(x-6)^2+4$. If this parabola is transformed left 4 units and up 4 units, what are the coordinates of the vertex of the transformed parabola?

$\textbf{(A) } (-10, 0) \qquad \textbf{(B) } (2, 8) \qquad \textbf{(C) } (6, 4) \qquad \textbf{(D) } (10, 8)$

\item Given the system of equations $y=x^2$ and $y=9x-20$, which of the following is a possible solution for $x$?

$\textbf{(A) } 2 \qquad \textbf{(B) } 4 \qquad \textbf{(C) } 6 \qquad \textbf{(D) } 8$

\item Which of the following shows the solutions for the equation $n^2-10n-24=0$?

$\textbf{(A) } -12 \text{ and } 2 \qquad \textbf{(B) } -6 \text{ and } 4 \qquad \textbf{(C) } 12 \text{ and } -2 \qquad \textbf{(D) } 6 \text{ and } -4$

\item How many real solutions does the function $k(x)$ have if $k(x)=11x^2+22x+11$?

\item If $x^2-px-14=0$ for all values of $x$ and the sum of the roots is 5, what is the value of $p$?
 
\item For $x>0$, what is one possible solution to the equation $x^3-x^2-9x+9=0$?
 
\item The function $f$ 9s defined as $f(x)=3x-5$. The function $g$ is defined as $g(x)=2f(x-2)+6$. The function $h$ is defined as $h(x)=-g(x)+3$. What is the value of $h(8)$?
 
\item The XYZ company produces widgets. The company uses the formula $p=-3w^2+kw+1200$ to model its profit, $p$, in dollars versus the number of widgets, $w$, sold. 
If XYZ Company makes \$1800 on sales of 10 widgets and $k$ is a constant, what is the maximum profit that XYZ Company can make for a given number of widgets sold? 

\end{enumerate}
\section*{Solutions}
\begin{enumerate}[label=\bfseries\arabic*.]

\item 14
\item D
\item A
\item 1
\item -21
\item 4
\item 3
\item 26
\item 9
\item D
\item 36
\item D
\item B
\item B
\item 16
\item D
\item 25
\item A
\item 2
\item 96
\item A
\item 13
\item 6.5
\item B
\item B
\item C
\item 1
\item 5
\item 1, 3
\item -29
\item 1875
\end{enumerate}
\end{document}